---
title: "Data report - Gapminder dataset"
author: "Saar Alon-Barkat"
date: "March 2024"
format:
  html:
    self-contained: true
    code-fold: true
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
---


Last update: `r Sys.time()`


```{r}
library(tidyverse)
library(gapminder)
library(gt)
library(gtsummary)
library(ggrepel)  
library(plotly)
library(sjPlot)
```



This report uses the Gapminder dataset, which includes real-world country-level panel data (1952-2007) for: 

* Life expectancy (`lifeExp`)

* Population size (`pop`)

* GDP percapita (`gdpPercap`)

The dataset is available from `gapminder` `r` package. For more information on gapminder see: https://www.gapminder.org/ 



```{r}
gapminder %>% 
  glimpse()
```

<br>

The following table summarizes the main variables across the years. 

```{r}
gapminder %>% 
  select(-country) %>% 
  tbl_summary(by = year)

```


# Life expectancy


```{r}
gapminder %>% 
      ggplot(aes(x=factor(year), y=lifeExp)) +
  geom_boxplot()+
stat_summary(fun.y = "mean", geom = "point",color="lightblue",size=3)+
  theme_classic() +
  labs(title = "Life expectancy increase over time",
       x = "",
       y = "Country life expectancy",
       caption = "Source:Gapminder") 
```

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
      ggplot(aes(x=lifeExp)) +
  geom_histogram(bins=12,fill="lightblue",color="white")+
  theme_classic() +
  guides(color = FALSE) +
  labs(title = "Life expectancy variation in 2007",
       x = "Country life expectancy",
       y = "Frequency",
       caption = "Source:Gapminder") 
  
```
```{r}
gapminder %>% 
      ggplot(aes(x=lifeExp)) +
  geom_histogram(bins=12,fill="lightblue",color="white")+
  theme_classic() +
  facet_wrap(vars(year))+
  guides(color = FALSE) +
  labs(title = "Life expectancy variation over time",
       x = "Country life expectancy",
       y = "Frequency",
       caption = "Source:Gapminder") 

```




```{r}
gapminder %>% 
      ggplot(aes(x=continent, y=lifeExp)) +
  geom_boxplot()+
stat_summary(fun.y = "mean", geom = "point",color="lightblue",size=3)+
  theme_classic() +
  labs(title = "Life expectancy across continents in 2007",
       x = "",
       y = "Country life expectancy",
       caption = "Source:Gapminder") 
```



```{r}
gapminder %>% 
  mutate(country_israel = ifelse(country=="Israel","Israel","")) %>% 
ggplot(aes(x=year, y=lifeExp, group=country, color = country_israel,size = country_israel)) +
    geom_line(alpha=0.5)+
  scale_colour_manual(name = "", values=c("lightblue","black"))+
    scale_size_manual(name = "", values=c(0.2,2))+
    theme_classic()+  
  labs(title = "Life expectancy increase across countries",
       x = "",
       y = "Country life expectancy",
       caption = "Source:Gapminder") 
```




```{r}
gapminder_lifeExp_tab <- gapminder %>%
    group_by(continent, year) %>%
    summarise(lifeExp=median(lifeExp))


gapminder_lifeExp_tab %>%
    ggplot(aes(x=year, y=lifeExp, color=continent)) +
     geom_line(size=1) + 
     geom_point(size=1.5) +
    geom_label_repel(data = gapminder_lifeExp_tab %>% filter(year == 2007), aes(label = continent),
                   nudge_x = 1,
                   nudge_y = 1) +
  theme_classic() +
  guides(color = FALSE) +
  labs(title = "Life expectancy increase across continents",
       x = "",
       y = "Country median life expectancy",
       caption = "Source:Gapminder") 
```

<br>



# GDP per capita


```{r}
gapminder %>% 
      ggplot(aes(x=factor(year), y=gdpPercap)) +
  geom_boxplot()+
stat_summary(fun.y = "mean", geom = "point",color="lightblue",size=3)+
  theme_classic() +
  labs(title = "GPD per capaita increase over time",
       x = "",
       y = "Country GPD per capaita",
       caption = "Source:Gapminder") 
```

You can see that there's an outlier  


```{r}
gapminder %>% 
  filter(year == 2007) %>% 
      ggplot(aes(x=gdpPercap)) +
  geom_histogram(bins=12,fill="lightblue",color="white")+
  theme_classic() +
  guides(color = FALSE) +
  labs(title = "Variation in GDP per capita 2007",
       x = "Country GPD per capaita",
       y = "Frequency",
       caption = "Source:Gapminder") 
  
```

```{r}
gapminder %>% 
      ggplot(aes(x=gdpPercap)) +
  geom_histogram(bins=12,fill="lightblue",color="white")+
  theme_classic() +
  facet_wrap(vars(year))+
  xlim(0,60000)+
  guides(color = FALSE) +
  labs(title = "Variation in GDP per capita over time",
       x = "Country GPD per capaita",
       y = "Frequency",
       caption = "Source:Gapminder") 

```


```{r}
gapminder %>% 
  filter(year==2007) %>% 
      ggplot(aes(x=continent, y=gdpPercap)) +
  geom_boxplot()+
stat_summary(fun.y = "mean", geom = "point",color="lightblue",size=3)+
  theme_classic() +
  labs(title = "GDP per capita across continents in 2007",
       x = "",
       y = "Country GPD per capaita",
       caption = "Source:Gapminder") 
```




```{r}
gapminder %>% 
  mutate(country_israel = ifelse(country=="Israel","Israel","")) %>% 
ggplot(aes(x=year, y=gdpPercap, group=country, color = country_israel,size = country_israel)) +
    geom_line(alpha=0.7)+
  scale_colour_manual(name = "", values=c("lightblue","black"))+
    scale_size_manual(name = "", values=c(0.2,2))+
    theme_classic()+  
  labs(title = "GPD per capaita increase across countries",
       x = "",
       y = "Country GPD per capaita",
       caption = "Source:Gapminder") 
```




```{r}
gapminder_gdpPercap_tab <- gapminder %>%
    group_by(continent, year) %>%
    summarise(gdpPercap=median(gdpPercap))


gapminder_gdpPercap_tab %>%
    ggplot(aes(x=year, y=gdpPercap, color=continent)) +
     geom_line(size=1) + 
     geom_point(size=1.5) +
    geom_label_repel(data = gapminder_gdpPercap_tab %>% filter(year == 2007), aes(label = continent),
                   nudge_x = 1,
                   nudge_y = 1) +
  theme_classic() +
  guides(color = FALSE) +
  labs(title = "GPD per capaita increase across continents",
       x = "",
       y = "Country GPD per capaita",
       caption = "Source:Gapminder") 
```

```{r}
gapminder_gdpPercap_tab <- gapminder %>%
  filter(country %in% c("Israel","Lebanon","Greece","Turkey","Syria")) %>% 
    group_by(country, year) %>%
    summarise(gdpPercap=median(gdpPercap))


gapminder_gdpPercap_tab %>%
    ggplot(aes(x=year, y=gdpPercap, color=country)) +
     geom_line(size=1) + 
     geom_point(size=1.5) +
    geom_label_repel(data = gapminder_gdpPercap_tab %>% filter(year == 2007), aes(label = country),
                   nudge_x = 1,
                   nudge_y = 1) +
  theme_classic() +
  guides(color = FALSE) +
  labs(title = "GPD per capaita increase across continents",
       x = "",
       y = "Country GPD per capaita",
       caption = "Source:Gapminder") 
```

# Link between life expectancy and GDP per capita



2007


```{r}
gapminder %>% 
  filter(year==2007) %>% 
  ggplot(aes(x=gdpPercap,y=lifeExp,color=continent,size=pop))+
  geom_point()+
    geom_text(aes(label=country),size=4,hjust=-0.1,vjust=-0.2,color="black",
            data = gapminder %>% filter(year==2007,country=="Israel"))+
    theme_classic()+
    guides(color = FALSE,size=FALSE) +
  labs(title = "Link between GDP per capita and life expectancy across continents",
       x = "GDP per capita",
       y = "Life expectancy",
       caption = "Source:Gapminder") 


```

```{r}


p1 <- gapminder %>% 
  filter(year==2007) %>% 
  ggplot(aes(x=gdpPercap,y=lifeExp,color=continent,size=pop, text = paste("country:", country)))+
  geom_point()+
    geom_text(aes(label=country),size=4,hjust=-0.1,vjust=-0.2,color="black",
            data = gapminder %>% filter(year==2007,country=="Israel"))+
    theme_classic()+
    guides(color = FALSE,size=FALSE) +
  labs(title = "Link between GDP per capita and life expectancy across continents",
       x = "GDP per capita",
       y = "Life expectancy",
       caption = "Source:Gapminder") 


ggplotly(p1)
```



```{r}
gapminder %>% 
  filter(year==2007) %>% 
  ggplot(aes(x=gdpPercap,y=lifeExp,color=continent,size=pop))+
  geom_point()+
  geom_smooth(aes(group=1))+
    geom_text(aes(label=country),size=4,hjust=-0.1,vjust=-0.2,color="black",
            data = gapminder %>% filter(year==2007,country=="Israel"))+
    theme_classic()+
    guides(color = FALSE,size=FALSE) +
  labs(title = "Link between GDP per capita and life expectancy across continents",
       x = "GDP per capita",
       y = "Life expectancy",
       caption = "Source:Gapminder") 


```




```{r}
gapminder %>% 
  filter(year==2007) %>% 
  ggplot(aes(x=log(gdpPercap),y=lifeExp,color=continent,size=pop))+
  geom_point()+
  geom_smooth(method="lm")+
    geom_text(aes(label=country),size=4,hjust=-0.1,vjust=-0.2,color="black",
            data = gapminder %>% filter(year==2007,country=="Israel"))+
    theme_classic()+
    guides(color = FALSE,size=FALSE) +
  labs(title = "Link between GDP per capita and life expectancy across continents",
       x = "GDP per capita (log)",
       y = "Life expectancy",
       caption = "Source:Gapminder") 


```



## Regression model

```{r}
tmod1 <- lm(lifeExp ~ log(gdpPercap) + year + continent, data=gapminder)


tab_model(tmod1)

plot_model(tmod1)
plot_model(tmod1,type = "std")

plot_model(tmod1,type = "pred")


```

```{r}
tmod2 <- tmod1 %>% update(.~. + continent* (log(gdpPercap) + year))


tab_model(tmod1,tmod2)

plot_model(tmod2,type = "pred", terms = c("year","continent"))

```



```{r}
anova(tmod1,tmod2)
```


